Variance reduction for Markov chains via martingale representations

### 2020-05-09 aap/Revision/Response_1.tex

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 1

\section*{Main remarks}

\begin{itemize}

\item \textit{... The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction. On these topics several results are presented as Theorems and Propositions  ... I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ... where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
\par
Let us stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms. Only few references studied MSE error of the Langevin algorithm (see e.g.  Durmus and  Moulines (2017)). In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms. The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-to-variance criterion. Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest. Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation. In  Section~4 we  explain in accessible way why our approach leads to reduced cost-to-variance criterion. These results are based on deep analysis of the variance reduced estimate postponed to Section 5. In the revised version we significantly weaken  assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).   Note that all assumptions needed are now collected in Theorem~3. For convenience of the reader we also sketch the main steps of the proof right after 
Theorem~3. We decided to present our main theoretical results as bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13). Given these estimates it is easy to see how to get bounds for the variance. 
\item \textit{... It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit. The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
\item \textit{... here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''. For many practitioners, $\alpha=0$ would be rather natural. What can be said in this important case? This is also applies to Section 6 ...}
\par
In the revised version we decided to switch to a constant time step and present all our bounds for this case. On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.  
\item \textit{ ... here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials. In particular the bound B would be equal to infinity. How can this be reconciled? ...}
\par
In the revised version we skip this result because of size restrictions. 
\end{itemize}

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-(1/2)\|x-\mu\|^2} + \rme^{-(1/2)\|x+\mu\|^2}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-variance ratio
\begin{eqnarray}
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]}
\end{eqnarray}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
 \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
 \end{equation*}
 since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$. Similarly, for $K = 2$ we set
  \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
 \end{equation*}
since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

@@

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig2}}
\caption{Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions.
\label{fig:gaus_cost}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig2}}
\caption{Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for a mixture of two-dimensional Gaussian distributions and different test functions.
\label{fig:gmm_cost}}
\end{figure}

>>>>

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-(1/2)\|x-\mu\|^2} + \rme^{-(1/2)\|x+\mu\|^2}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-to-variance ratio
\begin{eqnarray}
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]}
\end{eqnarray}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. Since for a fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to the cost of computing function $f$, we set for $K = 1$
 \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
 \end{equation*}
 since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}$ for a fixed $r$. Note that in this case each $a_{r,\mathbf{k}}$  is a linear function, which can be computed at the same cost as $f$. Similarly, for $K = 2$ we set
  \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8,
 \end{equation*}
since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for a fixed $r$. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

@@

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a two-dimensional standard Gaussian distribution and different test functions.
\label{fig:gaus_cost}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig2}}
\caption{Cost-to-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for the mixture \eqref{eq:gmm} of two-dimensional Gaussian distributions and two  test functions.
\label{fig:gmm_cost}}
\end{figure}
